---
title: "Setup"
---

This page helps you get R, packages, and the course repository running smoothly.

## 1) Install/update software

- **R** (recommended: recent CRAN release)
- **RStudio** (recommended)
- **Quarto** (needed only if you want to render slides/labs locally)

## 2) Get the course materials

### Option A — Git (recommended)
- Clone the repository
- Pull updates during the course if needed

### Option B — Download ZIP
- Download the ZIP from GitHub
- Unzip and open the project folder in RStudio

## 3) Install packages

Run this once (edit if you already have everything):

```{r}
pkgs <- c(
  "lavaan",
  "semTools",
  "psych",
  "MVN",
  "mice",
  "modsem",
  "ggplot2",
  "dplyr",
  "tidyr"
)

to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

invisible(lapply(pkgs, library, character.only = TRUE))
```

## 4) Reproducibility option (recommended): renv

If the repo uses `renv`, you can restore the exact package versions:

```{r}
# install.packages("renv")
# renv::restore()
```

## 5) Quick test: can you run lavaan?

```{r}
library(lavaan)

m <- 'f =~ x1 + x2 + x3'
fit <- cfa(m, data = HolzingerSwineford1939)

inspect(fit, "converged")
fitMeasures(fit, c("cfi","rmsea","srmr"))
```

## Troubleshooting

### “Package X is not available”
- Update R
- Try a different CRAN mirror
- Install from source if needed (rare)

### “Quarto not found”
- Install Quarto and restart RStudio

### “Rendering fails”
- Re-run package installation
- Try rendering a single file first (e.g., one lab)
- Check for path issues (especially on Windows)

::: {.callout-tip}
If you run into errors during class, copy-paste the **error message + the code you ran** into your notes (or send it). It speeds up debugging a lot.
:::
